---
title: "Robust_NLME"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/README-",
  out.width = "100%"
)
```

## This repo contains R code that was used for the real data analysis and simulations in my paper ``Jointly Modeling Means and Variances for Nonlinear Mixed Effects Models with Measurement Errors and Outliers''

-----------------------------------------------------

### What's in the repo

> `/src`: source code 

  + `00_dependencies.R`: the dependencies packages  
  + `Rnlme.R`: the function which jointly models mean and variance for NLME. Measurement error model can be added if needed.
  + `get_sd_bootstrap1.R`: the function produces adjusted SD by using parametric bootstrapping, when the random effect $a_i$ (see Model (2) in the paper) follows a normal distribution
  + `get_sd_bootstrap2.R`: the function produces adjusted SD by using parametric bootstrapping, when the random effect $a_i$ (see Model (2) in the paper) follows an inverse $\chi^2$ distribution
  + `get_Jloglike.R`, `est_fixed.R`, etc: all the other R files are self-defined functions sourced in the above functions and main analysis. 

> `/data`: 

  + `/toy data`: A toy data generated from simulation to illustrate how to use our methods. 
  + `raw_data.csv`: (**EMPTY**) DTP data used in the real analysis. The data that support the findings of this study are confidential due to privacy or ethical restrictions.

> `/simulation`: R files for different simulation settings (*see details about the simulation design in my paper Section 5*)

  + `Setting_I`: Setting I
  + `Setting_II_df=3`: Setting II with df=3
  + `Setting_II_df=5`: Setting II with df=5
  + `Setting_III`: Setting III (*The corresponding results shown in the Supplementery Materials*)
  + `Setting_IV`: Setting IV (*The corresponding results shown in the Supplementery Materials*)
  + `Setting_SUPP`: A supplementary simulation setting (*The corresponding results shown in the Supplementery Materials*)
  
> `/real data analysis`: R files for real data analysis (*see details in my paper Section 4*)

  + `00_dependencies.R`: the dependencies packages 
  + `01_preprocess.R`: reprocess the raw data and produce clean data for analysis
  + `02_model.R`:
  
> `/figures`: figures used in the paper based on the real data analysis
  
### How to re-produce the results in the paper
> Download the entire repo in your local path

> Simulations
  
  + Run each file in the `/simulation` folder


> Real data analysis
  
  + Run the following R scripts in the `\real data analysis` step by step
    + `00_dependencies.R`: install all dependencies packages
    + `01_data_process.R`: pre-process the raw data; make and save the analytic data
    + `02_model_fitting_adj_trt.R`: fit models for three methods: LB, TS, and JM

### Example: apply the proposed methods to the toy data

```{r example, eval=FALSE}

########################### load packages
library(nlme)
library(tidyverse)
library(Deriv)
library(stringr)
library(LaplacesDemon)
library(purrr)
library(MASS)
library(xtable)

rm(list=ls())
########################## source all functions  
(file.sources = list.files(path=here::here("src"),pattern="*.R$"))
(file.sources <- paste0(here::here("src"), "/", file.sources))
sapply(file.sources,source)

###############################################

#### Read in the data
simdat <- readRDS(here::here("data","toy_data.rds"))

# 100 unique patients
# 15 repeated measurements per patient
# var: patid, day, lgcopy, cd4

#### pre-define the functions we need for modeling
    nf1 <- function(p1,p2,p3,t) p1+p2*exp(-p3*t)
    nf2 <- function(p1,p2,p3,t) p1+p2*t+p3*t^2

#### LB method: nlme()
    cat("--Fit NLME() \n")
    dat_g <- groupedData(lgcopy~day|patid, data=simdat)
  
    
    start0 <- c(p1=1,p2=2,p3=3) # random starting values
    nls.fit  <- nls(lgcopy~nf1(p1,p2,p3, day), data=dat_g, start=start0)
    start <- coef(nls.fit) # update starting values
    nlme.fit <- withTimeout({try(nlme(lgcopy~nf1(p1,p2,p3, day),fixed = p1+p2+p3 ~1,random  = p1+p3 ~1,
                     data=dat_g,start=start))},timeout=1.5,onTimeout="warning")
#### TS method:
    cat("--Runing Two-step model\n")

    # step 1: fit CD4 model and get predicted true CD4 values
    
    cd4.fit <- try(lme(cd4~day+I(day^2), data=simdat, random=~1|patid))
    simdat$cd4.pred <- fitted(cd4.fit)
     
    # step 2: Joint modeling mean and variance for NLME
    
    # variance model:  
    sigmaObject_TS <- list(model=~1+cd4.pred+(1|patid),
                           link='log',
                           ran.dist="normal",
                           str.fixed=c(2*log(nlme.fit$sigma), 0),
                           lower.fixed=NULL,
                           upper.fixed=NULL,
                           fixName="alpha",
                           ranName="a",
                           dispName="siga",
                           str.disp=0.5)
    # mean model
    nlmeObject_TS <- list(nf = "nf1",
                          model= lgcopy ~ nf(p1,p2,p3,day),
                          var=c("day"),
                          fixed = p1+p2+p3 ~1,
                          random = p1+p3 ~1,
                          family='normal', 
                          ran.dist='normal',
                          fixName="beta",
                          ranName="u",
                          dispName="d",
                          sigma=sigmaObject_TS,    # residual dispersion model (include residual random eff)
                          ran.Cov=NULL,  # random effect dispersion model (include random random eff (double random eff))
                          str.fixed=fixef(nlme.fit),  # starting value for fixed effect
                          str.disp=as.numeric(VarCorr(nlme.fit)[,"StdDev"][c("p1", "p3")]),  # starting value for fixed dispersion of random eff
                          lower.fixed=NULL, # lower bounds for fixed eff
                          upper.fixed=rep(100,3), # upper bounds for fixed eff
                          lower.disp=c(0,0), # lower bounds for fixed dispersion of random eff
                          upper.disp=c(Inf,Inf) # upper bounds for  fixed dispersion of random eff)
                          )
    nlmeObjects_TS=list(nlmeObject_TS)
    
    TS <- try(Rnlme(nlmeObjects=nlmeObjects_TS, long.data=simdat, 
                        idVar="patid", sd.method="HL", dispersion.SD = TRUE,
                        independent.raneff=FALSE))

    
#### JM method
    cat("--Runing Joint model\n")
    
    #### JM
    sigma1 <- list(model=NULL,
                   str.disp=as.numeric(VarCorr(cd4.fit)[,"StdDev"][2]),
                   lower.disp=NULL,
                   upper.disp=NULL,
                   parName="xi")
    
    # measurement error model
    lmeObject_JM <- list(nf = "nf2" ,
                         model= cd4 ~ nf(p1,p2,p3,day),
                         var=c("day"),
                         fixed = p1+p2+p3 ~1,
                         random = p1 ~1,
                         family='normal', 
                         ran.dist='normal',
                         fixName="gamma",
                         ranName="b",
                         dispName="sigb",
                         sigma=sigma1,    # residual dispersion model (include residual random eff)
                         ran.Cov=NULL,  # random effect dispersion model (include random random eff (double random eff))
                         str.fixed=fixef(cd4.fit),  # starting value for fixed effect
                         str.disp=as.numeric(VarCorr(cd4.fit)[,"StdDev"][1]),  # starting value for fixed dispersion of random eff
                         lower.fixed=NULL, # lower bounds for fixed eff
                         upper.fixed=rep(100,3), # upper bounds for fixed eff
                         lower.disp=c(0), # lower bounds for fixed dispersion of random eff
                         upper.disp=c(Inf) # upper bounds for  fixed dispersion of random eff
                         )
    # variance model:  
    sigma2 <- list(model=~1+cd4.true+(1|patid),
                   link='log',
                   ran.dist="normal",
                   str.fixed=c(2*log(nlme.fit$sigma), 0),
                   lower.fixed=NULL,
                   upper.fixed=NULL,
                   fixName="alpha",
                   ranName="a",
                   dispName="siga",
                   str.disp=0.5,
                   trueVal.model=list(var="cd4.true", model=lmeObject_JM)
                   )
    
    # mean model
    nlmeObject_JM <- list(nf = "nf1",
                          model= lgcopy ~ nf(p1,p2,p3,day),
                          var=c("day"),
                          fixed = p1+p2+p3 ~1,
                          random = p1+p3 ~1,
                          family='normal', 
                          ran.dist='normal',
                          fixName="beta",
                          ranName="u",
                          dispName="d",
                          sigma=sigma2,    # residual dispersion model (include residual random eff)
                          ran.Cov=NULL,  # random effect dispersion model (include random random eff (double random eff)
                          str.fixed=fixef(nlme.fit),  # starting value for fixed effect
                          str.disp=as.numeric(VarCorr(nlme.fit)[,"StdDev"][c("p1", "p3")]),  # starting value for fixed dispersion of random eff
                          lower.fixed=NULL, # lower bounds for fixed eff
                          upper.fixed=rep(100,3), # upper bounds for fixed eff
                          lower.disp=c(0,0), # lower bounds for fixed dispersion of random eff
                          upper.disp=c(Inf,Inf) # upper bounds for  fixed dispersion of random eff
                          )
    
    nlmeObjects_JM <- list(nlmeObject_JM, lmeObject_JM)
    
    JM <- try(Rnlme(nlmeObjects=nlmeObjects_JM, long.data=simdat, 
                    idVar="patid", sd.method="HL", dispersion.SD = TRUE,
                    independent.raneff="byModel"))
```
  
