---
title: "get_info_sigma()"
author: "Qian Ye"
date: "2026-06-30"
output: html_document
header-includes:
  - \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This function is used to extract information from a pre-defined object. This object specifies a model for the within-individual variance for a longitudinal NLME model (called Variance model).

### *The NLME mean model*
For example, we consider the following NLME model 
$$
     \boldsymbol{y}_i = g(X_i, \boldsymbol{\beta}, \boldsymbol{u}_i) + \boldsymbol{e}_i, \qquad   i=1, \cdots, n, 
$$

where $g(\cdot)$ is a known nonlinear function for the mean $E(\boldsymbol{y}_i)$ of the response, matrix $X_i$ contains covariates, vector $\boldsymbol{\beta}$ contains fixed mean parameters, vector $\boldsymbol{u}_i$ contains random effects, and $\boldsymbol{e}_i = (e_{i1}, e_{i2}, \cdots, e_{in_i})^T$ are within-individual random errors.

We assume that 
$$ 
e_{ij} \sim N(0, \sigma_{ij}^2), \quad  i=1,2,\cdots,n, \; j=1,2,\cdots,n_i. 
$$
Note that the variance of the within-individual random errors, $\sigma_{ij}^2$, is allowed to be individual-specific and time-dependent. 

---

### *The Variance model*

We consider the following model for the within-individual variance 
$$
     \log(\sigma_{ij}^2) = \phi(\boldsymbol{\alpha}, z_{ij}) +  a_i,  \qquad i=1, \cdots, n,\, j=1, \cdots, n_i, 
$$
where $\phi(\cdot)$ is a known function, $\boldsymbol{\alpha}$ contains fixed parameters, and $a_i$ is a random effect for the variance model.

---

### *How to define the object for the variance model*

When define variance model, we can assume the covariates in the model are either observed or latent.


1. The following code shows the example that the covariate $z$ are observed.


```{r,results='hide', message=FALSE}
# read in all the source code files
(file.sources = list.files(path=here::here("src"),pattern="*.R$"))
(file.sources <- paste0(here::here("src"), "/", file.sources))
sapply(file.sources,source)

# choose starting values
alpha0 <- -8
alpha1 <-  1.5
sigma_a <- 0.7

# define the variance model
sigma <- list(
  model=~1+z+(1|patid), # specify the model structure: a linear model with a intercept random effect a_i
  link='log',           # specify the link function
  ran.dist="normal",    # specify the distribution for the random effect a_i 
                        # options are: "inverse-Chi" or "normal" or "stdnormal"
  str.fixed=c(alpha0, alpha1), # specify the starting values for the fixed effects
  lower.fixed=NULL,     # lower limits for the fixed effects
  upper.fixed=NULL,     # upper limits for the fixed effects
  fixName="alpha",      # specify the notations for fixed effects
  ranName="a",          # specify the notations for random effects
  dispName="siga",      # specify the notations for the dispersion parameters for the random effects
  str.disp=sigma_a      # specify the starting values for the dispersion parameters for the random effects
)

```

Then we can use `get_info_sigma()` to extract the information from the variance model. The output values contain:

| Values | Description |
|:-------|:------------|
|`sigmaExpr`| formula for the variance $\sigma_{ij}$ |
|`raneff`| random effects |
|`fixed`| fixed effects |
|`df`| degree of freedom of the inverse-chi distribution, if applicable |
|`str.val`| starting values for the fixed effects |
|`lower`| lower limits for the fixed effects |
|`upper`| upper limits for the fixed effects |
|`disp`| the dispersion parameters for the random effects |
|`str.disp`| starting values for the dispersion parameters|
|`lower.disp`| lower limits for the dispersion parameters |
|`upper.disp`| upper limits for the dispersion parameters |

```{r }
get_info_sigma(sigma)


```

2. The following code shows the example that the covariate $z$ are latent, and its true values are specified by a measurement error model.

First, we need to define the measurement error model
```{r, results='hide', message=FALSE}
# define a linear function for the measurement error model
nf <- function(p1,p2,p3,t) p1+p2*t+p3*t^2

# choose starting values
gamma <- c(5.2, 1.6, -1.2)
xi <- 0.2
sigma_b <- 0.4

# define a pure random error term for the measurement error model
sigma1 <- list(
  model=NULL,
  str.disp=xi,
  lower.disp=NULL,
  upper.disp=NULL,
  parName="xi"
)

# define the LME measurement error model
lme_MEM <- list(
  nf = "nf" ,
  model= z ~ nf(p1,p2,p3,day), # specify the model structure
  var=c("day"),                # specify all the variables used in the model
  fixed = p1+p2+p3 ~1,         # specify the fixed effects
  random = p1 ~1,              # specify the random effects
  family='normal',             # specify the distribution for the error term
  ran.dist='normal',           # specify the distribution for the random effects
  fixName="gamma",             # specify the notations for the fixed effects
  ranName="b",                 # specify the notations for the random effects
  dispName="sigb",             # specify the notations for the dispersion parameters for the random effects
  sigma=sigma1,                # specify the pre-defined the error term
  ran.Cov=NULL,           
  str.fixed=gamma,             # specify the starting value for fixed effects
  str.disp=c(sigma_b),         # specify the starting value for the dispersion parameters for the random effects
  lower.fixed=NULL,            # specify the lower bounds for fixed effects
  upper.fixed=rep(100,3),      # specify the upper bounds for fixed effests
  lower.disp=c(0),             # specify the lower bounds for dispersion parameters
  upper.disp=c(Inf)            # specify the upper bounds for dispersion parameters
)

```
Then we define the variance model with latent covariate $z$
```{r}
  
sigma <- list(
  model=~1+z+(1|patid),
  link='log',
  ran.dist="normal",
  str.fixed=c(alpha0, alpha1),
  lower.fixed=NULL,
  upper.fixed=NULL,
  fixName="alpha",
  ranName="a",
  dispName="siga",
  str.disp=sigma_a,
  trueVal.model=list(var="z", model=lme_MEM) # specify the meassurement error model for the latent z
)
```

Finally, we can use `get_info_sigma()` to extract the information from the variance model.

```{r}
get_info_sigma(sigma)

```